---
title: "Assignment 5: Mixed models"
author: "Marton Kovacs / Zoltan Kekecs"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction

This assignment is related to the previous lab assignment concerning perioperative pain and its psychological and hormonal predictors. Just like previously, you will set up linear model to predict postoperative pain after wisdom tooth surgery, but this time you will have to also take into account the fact that there is clustering in the data. 

Your research paper on the effect of the psychological and hormonal predictors of postoperative pain was so successful, that you managed to secure research funding for a multi-site replication study. Here your collaborators collect data in the same way you did in the original study at 20 different hospital sites. The goal of the study is to increase the generalizability of your findings. You would like to assess the model coefficients and the overall predictive efficiency of the predictors in your model.

As a reminder, here is the protocol for data collection: “You have collected data from adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset. 

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgemental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.”

# Datasets

You will need two datasets for this assignment, datafile A and B. You can load them from the 'data/' folder.

# Task

First, fit a linear mixed model to estimate postoperative pain on datafile A. You should use the same fixed effect predictors as you used in your final model in the 3 assignment. (If you did not do that assignment, use the following predictors: age, sex, STAI, pain catastrophizing, mindfulness, and serum cortisol.) Importantly, the model has to account for the clustering of the data in different hospital sites. We have no reason to assume that the effects of the different predictors would be different in the different hospitals, so fit a random intercept model including the random intercept of hospital-ID. Once the model is built, note the model coefficients and the confidence intervals of the coefficients for all fixed effect predictors, and compare them to the ones obtained in the 3 assignment. 

Also, compute the variance explained by the fixed effect predictors using marginal R^2^, and the variance explained by the fixed and random effect terms combined using conditional R^2^. Now use the model coefficients obtained on data file A to predict pain in datafile B.

__IMPORTANT:__ Do not fit the regression models on data file B (don’t re-train your models), just use the regression equation you derived based on datafile A. These regression equations should be applied on the new data (datafile B), to predict pain.

Now compute the variance explained by the model on datafile B. You can do this by using the formula: __1 - (RSS / TSS) = R^2__. Compare this R^2^ to the marginal and conditional R^2^ values computed for the model on datafile A. 

# What to report

Report the model coefficients and the confidence intervals of the coefficients for each fixed effect predictor obtained on data file A in a table. 

Report the variance components for the fixed effects, the random intercept, and the residuals (from the model on data file A). Also report the marginal R^2^ and the conditional R^2^ squared obtained from the model on data file A, and the observed R^2^ of this model for data file B.

# What to discuss

Compare the model coefficients and the confidence intervals observed in this assignment and the assignment for Lab 2 and discuss what you think the differences or similarities mean.

# Solution

## Read the data

Read the datasets used in this assignment. Pay attention to the extensions of the datafiles.

## Load libraries
```{r}
if(!require("haven")) install.packages("haven") # package that is used when reading .sav datafiles
library(haven)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(lmerTest)
library(lme4)
```


# ===================================================================================================
# DATASET A - EXPLORATORY DATA ANALYSIS

```{r}
# Dataset A: 
a_raw_data <- read_spss("Documents/2025_26_2/r-course-Spring-2026/data/assignment_4_dataset_A.sav")
```

## Exploratory data analysis

Run an exploratory data analysis (EDA) to investigate the dataset.

### 1. Data structure
The first step was to investigate the structure of the dataset: this included the types and ranges of variables, and the potential presence of missing values.

#### Data types
```{r}
str(a_raw_data) 
```
The variables are mostly numerical (pain, age, STAI_trait, pain_cat, cortisol_serum, cortisol_saliva, mindfulness, weight, IQ, household_income), while strings can also be found in the dataset (ID, sex, hospital). 

#### Missing values
```{r}
colSums(is.na(a_raw_data))
```
No missing data was found within the variables. 

#### Value ranges 
```{r}
summary(a_raw_data)
```
All the variables were in synch with the task description, the ranges were identical to the ones mentioned above. 

### 2. Distributions
#### 2.1. Distribution of postoperative pain scores (`pain`)
```{r}
ggplot(a_raw_data, aes(x = pain)) +
  geom_histogram() +
  labs(
    title = "Distribution of prosperative pain scores"
  )
```
The scores appeared to be approximately symmetrically dirstibuted. The distribution shows no strong skewness, and can be considered approximately normal.

#### 2.2. Psychological predictors
Next, the distributions of the following psychological variables were examined: 
* `STAI_trait` (trait anxiety), 
* `pain_cat` (pain catastrophizing), 
* `mindfulness`.

##### Trait anxiety - `STAI_trait`
```{r}
range(a_raw_data$STAI_trait)
# Range is between 24 and 52. 
# The variable measures anxiety on a scale of 20-80.

ggplot(a_raw_data, aes(x = STAI_trait)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(20, 60)) +
  labs(title = "Distribution of trait anxiety")

# The theoretical maximum limit of the scale are not displayed on the axis. 
```
The distribution appears to have a slight right skew - indicating more observations at higher levels of anxiety. No extreme outliers were observed. The maximum observed value was 52, considerably below the theoretical upper bound of the scale (80). This suggests that very high levels of trait anxiety were not present in this sample. 

##### Pain catastrophizing - `pain_cat`
```{r}
range(a_raw_data$pain_cat)
# Range is between 17 and 42. Theoretical minimum on the scale is 0, the theoretical maximum is 52. 

ggplot(a_raw_data, aes(x = pain_cat)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 52)) +
  labs(title = "Distribution of pain catastrophizing")

# The x axis limits reflect the theoretical range of the pain catastrophizing scale. 
```
The distribution show that majority of the observations are concentrated between approximately 25 and 35, indicating an overally moderate level of catastrophizing.


##### Mindfulness
```{r}
range(a_raw_data$mindfulness)
# Range is between 1.00 and 5.58, which is in connection with the theoretical range of 1-6.

ggplot(a_raw_data, aes(x = mindfulness)) +
  geom_histogram(aes(y = after_stat(density)), 
                 binwidth = 0.1, 
                 fill = "#2f9147") +
  geom_density() +
  labs(title = "Distribution of dispositional mindfulness")
```
The distribution of dispositional mindfulness was visualized using a histogram. In order to show the shape of the distribution more smoothly, a density line was added to the histogram. 
Most observations are concentrated between approximately 2.5 and 4, indicating mild levels of dispositional mindfulness in the sample. Considerably high values are relatively infrequent, no clear extreme outliers can be detected. 


#### 2.3. Hormonal variables - cortisol
The distributions of `cortisol_serum` and `cortisol_saliva` were also visualized  with a faceted histograms and density curves.

```{r}
cortisol_data <- a_raw_data %>% 
  select(cortisol_serum, cortisol_saliva) %>% 
  pivot_longer(cols = everything(), 
               names_to = "type", 
               values_to = "cortisol")
```
Here, the data was formatted in order to create the plot. 

```{r}
ggplot(cortisol_data, aes(x = cortisol)) +
  geom_histogram(aes(y = after_stat(density)), 
                 binwidth = 0.15, 
                 fill = "#328da8") +
  geom_density() +
  facet_wrap(~ type) +
  labs(title = "Distribution of cortisol levels")
```
The distributions show similar shape. However, the serum cortisol values seem to be centered at a slightly higher level, compared to salivary cortisol.

#### 2.4. Demographic variables 
Finally, the distributions of the following demographic variables were visualized: 
* age,
* sex,
* and weight.

```{r}
age_sex_data <- a_raw_data %>% 
  select(age, sex) %>% 
  mutate(age = as.factor(age)) %>% 
  mutate(sex = recode(sex, "Male" = "male"))

age_pyramid(age_sex_data, age_group = age, split_by = sex) +
      labs(x = "Age group", 
          y = "Number of people", 
          title = "Distribution of age within the genders")
```
The age distribution of participants was visualized separately for female and male participants. Ages ranged approximately from 28 to 51 years. The distributions for males and females appear to be similar, no clear systematic difference can be observed.

Visualization of weight within males and females: 
```{r}
sex_weight_data <- a_raw_data %>% 
  select(weight, sex) %>% 
  mutate(sex = recode(sex, "Male" = "male"))

ggplot(sex_weight_data, aes(x = sex, y = weight)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, 
              alpha = 0.4)
```

### 3. Bivariate relationships - pain and predictors
#### 3.1. Pain ~ STAI_trait 
```{r}
stai_pain_data <- a_raw_data %>%
  group_by(pain) %>%
  summarise(mean_STAI = mean(STAI_trait),
            se = sd(STAI_trait)/sqrt(n())) 
  ggplot(stai_pain_data, aes(x = pain, y = mean_STAI)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean_STAI - se,
                    ymax = mean_STAI + se))
```
+ interpretation

#### 3.2. Pain ~ pain_cat 
```{r}

```


#### 3.3. Pain ~ mindfulness 
#### 3.4. Pain ~ cortisol_serum
#### 3.5. Pain ~ cortisol_saliva 




## Relationships among predictors

## Do sites differ? - Clustering structure

Before modeling, search answers for: 
- should I transform (log)?
- should i stardardize predictors? 

## =============================================================
## Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}

```

