---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuazlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTuesday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
library(tidyverse)
library(dplyr)
library(tidytuesdayR)
library(ggplot2)
```


### Loading data
```{r}
tuesdata <- tidytuesdayR::tt_load('2020-09-22') 
expeditions <- tuesdata$expeditions
```

### Aggregating the data
- with the usage of `forcats::fct_lump` and `fct_reorder`
```{r}
library(forcats)
expeditions_prep <- expeditions %>% 
  select(
    peak_name, season
  ) %>% 
  mutate(
    peak_name = fct_lump(peak_name, n = 15)) %>% 
  filter(peak_name != "Other") %>% 
  mutate(
    peak_name = fct_reorder(peak_name, peak_name, length)
  )
```

### Creating the plot
```{r}
library(viridis)
expeditions_plot <- ggplot(expeditions_prep, aes(y = peak_name)) +
  geom_bar(aes(fill = season), position = position_stack()) +
  scale_fill_viridis_d(option = "viridis") +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    x = "Number of expeditions",
    y = NULL
  )

expeditions_plot
```


## ======================================================
## Task 2: PhDs awarded

The 2019-02-19 TidyTuesday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

### Loading data
```{r}
phd_by_field <- tidytuesdayR::tt_load('2019-02-19')
phd_data <- phd_by_field$phd_by_field
```

### Aggregating data
```{r}
phd_processed <- phd_data %>% 
  select(
    year, broad_field, n_phds
  ) %>%
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(n_phds = sum(n_phds), .groups = "drop")
```


### Creating the plot
```{r}
library(scales)
phd_plot <- ggplot(phd_processed, aes(x = year, y = n_phds, colour = broad_field)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "Dark2", 
                     name = "Broad field") +
  labs(
    title = "Number of awarded Ph.D.-s in the US by year",
    x = NULL,
    y = NULL
  ) +
  scale_x_continuous(breaks = pretty_breaks()) + ## x axis formatting
  scale_y_continuous(labels = comma_format()) ## y axis formatting 

phd_plot
```


## ======================================================
## Task 3: Commute in the US

The 2019-11-05 TidyTuseday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

## Importing data
```{r}
commute_us <- tidytuesdayR::tt_load('2019-11-05')
commute_data <- commute_us$commute
```

## Aggregating data
```{r}
commute_processed <- commute_data %>% 
  select(n, mode, state_abb, state_region) %>% 
  filter(!is.na(n)) %>% 
  group_by(state_abb, mode, state_region) %>% 
  summarise(n = sum(n), .groups = "drop") %>% 
  pivot_wider(
    names_from = mode,
    values_from = n
  ) %>% 
  filter(!(is.na(state_abb))) %>% 
  rename(walk = Walk, 
         bike = Bike)
  
```

## Creating the plot 
```{r}
commute_plot <- ggplot(commute_processed, aes(x = walk, y = bike, colour = state_region)) +
  geom_point(aes(colour = state_region)) +
  labs(
    title = "Title of people walking vs. biking to work in each USA state",
    x = "Number of ppl walking to work (log N)",
    y = "Number of ppl biking to work (log N)"
  ) +
  scale_y_log10(labels = comma_format()) +
  scale_x_log10(labels = comma_format()) +
  geom_text(aes(label = state_abb), colour = "black")
  

## TODO: geom_text labels -> too crowded, also legend title

commute_plot
```

